Self Referencing Sequences

### 2014-11-08 draft.tex
heavily modified substitution rules section

Before we do so however we will clarify two different notions of convergence, that will allow us to state our result more precisely. 
\begin{definition} We say that a sequence $s^{(i)}$ is weakly convergent to $s(C,a)$ if $s^{\infty}$ exists and $s^{\infty} = s(C,a)$. We say that the sequence $s^{(i)}$ is strongly convergent to $s(C,a)$ if $s^{(i)}$ is a prefix of $s(C,a)$ for all $i \geq k$ for some $k\geq 0$. 
\end{definition}
Note that all strongly convergent sequences are also weakly convergent. We now can show that if we specify a generating cycle any one element starting sequence (other than $1$) is strongly convergent to the self referencing sequence that begins with the element.

\begin{theorem}
\label{iteratetheorem}
Let $C$ be some generating cycle. Let $s^{(0)} = a$ for some $a \in C$ where $a \neq 1$, and let $s^{(i)}$ be the sequences generated by the iterative process. Then $s^{(i)}$ is strongly convergent to $s(C,a)$.
\end{theorem}

\begin{proof}
We prove this by induction on $i$. We show that if $s^{(i)}$ is a prefix of $s(C,a)$, then $s^{(i+1)}$ is a longer prefix of $s(C,a)$. The base case of the induction is satisfied, as $s^{(0)}$ is a prefix of $s(C,a)$.

For the inductive step, we observe that a prefix of the sequence describes the block lengths of a \emph{longer} prefix of the sequence. We know that the sequence it describes is longer because the length of $s^{(i+1)}$ equals the sum of of the numbers in $s^{(i)}$, which is strictly greater than the length of $s^{(i)}$.

%The resultant sequence is a prefix of the self referencing sequence because if we apply our process to any prefix of the self referencing sequence it will produce a prefix of the self referencing sequence.

Since very $s^{(i)}$ is a prefix of $s(C,a)$, we know that once $s^{(i)}_{j}$ has been assigned a value, $s^{(k)}_{j}$ will have the same value for all $k \geq i$. Because of this the limit is well defined and so $s^{\infty}$ exists. As the limit of the process we know that $s^{\infty}$ is a prefix of $s(C,a)$, but since at every step of the process the length of the sequence strictly increases we must have that $s^{\infty}$ has infinite length, and so $s^{\infty} = s(C,a)$. 
\end{proof}

Note that we must specify that $a \neq 1$, otherwise $s^{(i)}$ is always $1$, and so the length of the sequence does not grow.

Above we only considered starting sequences which have length one,  but we would like to consider arbitrary starting sequences. Going back to the case of $C = (1 2)$, we can let $s^{(0)}$ be \emph{any} sequence of 1's and 2's that starts with a 2 and the limiting sequence will still be the self-referential sequence.
For example, suppose $s^{(0)} = 2,1,1,1,2$. The rules for generating $s^{(i)}$ give:
\begin{align*}
s^{(0)}&= \mathbf{2},1,1,1,2\\
s^{(1)} &= \mathbf{2,2,1},2,1,2,2\\
s^{(2)} &= \mathbf{2,2,1,1,2,1},1,2,1,1,2,2\\
s^{(3)} &= \mathbf{2,2,1,1,2,1,2,2,1,2},1,1,2,1,2,2,1,1\\
&\vdots
\end{align*}
Here, we bold the numbers that form a prefix of $s(C,2)$. We note that $s^\infty = s(C,a)$ because $s^{(0)}$ contains a 2 at the beginning, which by the above theorem will produce $s(C,a)$ in the limit. This is true in general.

\begin{corollary}
Let $C$ be some generating cycle, let $s^{(0)}$ be any sequence of elements of $C$ which begins with some $a \in C$ such that $a \neq 1$. Then $s^{(i)}$ is weakly convergent to $s(C,a)$. 
\end{corollary}
\begin{proof} 
This follows directly from Theorem \ref{iteratetheorem}. That is because $s^{(0)}$ begins with an $a$, at each step of the process $s^{(i)}$ will be a prefix of $s(C,a)$ followed by some terms that disagree with $s(C,a)$. In the limit the prefix of $s(C,a)$ becomes $s(C,a)$ and so $s^{\infty} = s(C,a)$.
\end{proof}
Note that we do not guarantee strong convergence as there are some starting sequences, such as $2222$ which in numerical simulation appears to produce sequences which disagree with the $s( (12), 2)$ on a constant fraction of indices. We would like to be able to produce a general result that distinguishes between strongly and weekly convergent subsequences but it is unclear that there is a nice categorization of them.

@@

The iterative process given in Section~\ref{iterativeprocess} is one way to generate self-referencing sequences. However, for certain generating cycles $C$, there is a simpler ways to perform this iteration using a fixed set of substitution rules. Take $C = (1 3)$ as an example, and consider the sequence that starts with a 3:
\begin{align*}
333111333131333111333\dots
\end{align*}
We can use the iterative process from Section~\ref{iterativeprocess}, and suppose that $s^{(0)} = 33$. We get:
\begin{align*}
\{s^{(0)}\} &= 3,3\\
\{s^{(1)}\} &= 3,3,3,1,1,1\\
\{s^{(2)}\} &= 3,3,3,1,1,1,3,3,3,1,3,1\\
\{s^{(3)}\} &= 3,3,3,1,1,1,3,3,3,1,3,1,3,3,3,1,1,1,3,3,3,1,3,3,3,1\\
&\cdots
\end{align*}
It turns out that we can model this evolution with the following substitution rules:
\begin{align*}
\text{Rule }A: 3,3 &\to 3,3,3,1,1,1 \\
\text{Rule }B: 3,1 &\to 3,3,3,1\\
\text{Rule }C: 1,1 &\to 3,1
\end{align*}
This is to be interpreted as follows. Starting from 3,3, the only possible rule to apply is Rule A, which gives $3,3,3,1,1,1$ at the following iteration. We then apply Rule A to $3,3$, Rule B to $3,1$ and Rule C to $1,1$. Note that at every step, we break up the sequence into chunks of 2 and then simultaneously apply \emph{all} the rules. This will always be possible, since the rules preserve the fact that these sequences have even length. So at the following step, we have 3,3,3,1,1,1,3,3,3,1,3,1, which is $s^{(2)}$. Notice that in general, after the $i$th application of the substitution rules, the sequence is equal to $s^{(i)}$.

\begin{comment}
Why does this work? Well, first of all, notice that each rule of the form $x \rightarrow y$ has the . Also, note that we have decided to make the rules of length 2; if we had made the self-referential rules $1 \rightarrow 1$ and $3 \rightarrow 333$, this would not have worked. In addition, we know that each block length preserves parity.

We would like to know if we can do this for any generating cycle $C$ and starting number $a$. In particular we would like to be able to simulate our iterative process with substitution rules.

When we apply our substitution rules we should not need to know what happened to blocks other than the one we are currently substituting for. Therefore, each rule must somehow encode the length of  blocks, the number of elements already written in the current block, and what number the current block is composed of. The first condition is simple to implement, as the rules will specify the length of blocks. In order to keep track of the numbers already written, each rule should be self-contained. That is each rule should write complete blocks. In order to comply with the third condition it is enough for the size of the generating cycle to divide the length of the rules. This is so that at the start of each rule, we always know which numbers to write.
\end{comment}

We can formulate a set of sufficient conditions for rules that generate the self-referencing sequences. Formally, we will set $R$ to be a rule set, which is a set of strings of the form $x \rightarrow y$.

\begin{proposition}
\label{ruleprop}
Suppose $C$ is a cycle. For any $l \in \mathbb{N}$ divisible by $|C|$ and $a \in C$, let $R$ be a rule set. If for each string $s$ of elements in $C$ of length $l$, there exists a unique rule $x \rightarrow y$ in $R$ with $x = s$ and:
\begin{enumerate}
\item $x$ reads the block lengths of $y$ from left to right
\item $y$ starts with $a$
\item $|y|$ is divisible by $l$
\end{enumerate}
Then sucessively applying the rules in $R$, starting with a sequence of length $l$ generates a sequence which is weakly convergent to $s(C, a)$.
\end{proposition}

\begin{proof}
Let $s^{(i)}$ be the sequence generated from the process from Section~\ref{iterativeprocess}. I will show by induction on $i$ that $s^{(i)}$ is the result of applying the rules in $R$ $i$ times to a starting sequence $s^{(0)}$ of length $l$. By letting the starting sequence be $s^{(0)}$, we know it works for $i = 0$.

Suppose that applying the rules in $R$ to $s^{(0)}$ $i$ times gives $s^{(i)}$. Note that $|s^{(i)}|$ is divisible by $l$ since each iteration applies the rule to a segment of length $l$ and by condition 3, it writes a sequence of length a multiple of $l$. So we can partition $s^{(i)}$ uniquely into disjoint chunks of size $l$.

In the iterative process, each chunks of size $l$ will generate $l$ blocks, and since $|C|$ divides $l$, the block referenced by each of these chunks will have the same number. If the rules begin with $a$, then this number will be $a$.

So applying the rules to each chunk is equivalent to applying the iterative process, where we  group elements into chunks of size $l$. Therefore, $s^{(i+1)}$ is the result of applying the rules to $s^{(i)}$.
\end{proof}

Given Proposition~\ref{ruleprop} we can generate rules a large class of generating cycles $C$. Take the following example, where $C = (2 4)$, and the rules are
\begin{align*}
\text{Rule }A: 2,2 &\to 2, 2, 4, 4 \\
\text{Rule }B: 2,4 &\to 2, 2, 4,4,4,4\\
\text{Rule }C: 4,2 &\to 2,2,2,2,4,4\\
\text{Rule }D: 4,4 &\to 2, 2, 2, 2, 4, 4, 4, 4
\end{align*}
These rules satisfy the conditions of Proposition~\ref{ruleprop}, and so successive application of these rules will generate a self-referencing sequence which is weakly convergent to $s(C, 2)$.

@@

\begin{proof}
Suppose $l$ is a positive integer. Then $l$ must be even. Let $s$ be the string with one 2 and $l-1$ 1's. Suppose $x \rightarrow y$ is the rule for $s$, so $x = s$. Then by condition 1 of Proposition~\ref{ruleprop}, $|y| = l+1$, and by condition 3, $l = 1$. Which contradicts the fact that $l$ is divisible by $|C| = 2$. 
\end{proof}

@@

An open question regarding the self-referential sequence of 1's and 2's, $s((1 2),1)$, is what the limiting \emph{density} of 1's and 2's is. In this section, we consider the related problem of finding the density for sequences with generating cycle $C = (1 3)$ (note that the sequence that starts with a 1 differs by only one number from the sequence that starts with a 3, so we do not need to consider them separately). While we cannot prove that a limiting density exists, we give a numerical calculation of what the density is \emph{if it exists}. We use a general technique that takes advantage of the fact that substitution rules when $C = (1 3)$, as shown in Section~\ref{substitutionrules}.

>>>>

Before we do so however we will give two different notions of convergence that will allow us to state our result more precisely. 
\begin{definition} We say that a sequence $s^{(i)}$ is weakly convergent to $s(C,a)$ if $s^{\infty}$ exists and $s^{\infty} = s(C,a)$. We say that the sequence $s^{(i)}$ is strongly convergent to $s(C,a)$ if $s^{(i)}$ is a prefix of $s(C,a)$ for all $i \geq k$ for some $k\geq 0$. 
\end{definition}
Note that all strongly convergent sequences are also weakly convergent. We can now show that if we specify a generating cycle, any length 1 starting sequence $s^0$ (other than $1$) is strongly convergent to the self referencing sequence that begins with the element.

\begin{proposition}
\label{iteratetheorem}
Let $C$ be some generating cycle. Let $s^{(0)} = a$ for some $a \in C$ where $a \neq 1$, and let $s^{(i)}$ be the sequences generated by the iterative process. Then $s^{(i)}$ is strongly convergent to $s(C,a)$.
\end{proposition}

\begin{proof}
This claim is straightforward, but we will formally prove it with induction on $i$. We show that if $s^{(i)}$ is a prefix of $s(C,a)$, then $s^{(i+1)}$ is a longer prefix of $s(C,a)$. The base case of the induction is satisfied, as $s^{(0)}$ is a prefix of $s(C,a)$.

For the inductive step, we observe that a prefix of the sequence describes the block lengths of a \emph{longer} prefix of the sequence. We know that the sequence it describes is longer because the length of $s^{(i+1)}$ equals the sum of of the numbers in $s^{(i)}$, which is strictly greater than the length of $s^{(i)}$. We know that the sequence it describes is a prefix, since the first number matches up with $s(C,a)$, and so do the block lengths. This completes the induction.
%
%%The resultant sequence is a prefix of the self referencing sequence because if we apply our process to any prefix of the self referencing sequence it will produce a prefix of the self referencing sequence. 
%
%Since very $s^{(i)}$ is a prefix of $s(C,a)$, we know that once $s^{(i)}_{j}$ has been assigned a value, $s^{(k)}_{j}$ will have the same value for all $k \geq i$. Because of this the limit is well defined and so $s^{\infty}$ exists. As the limit of the process we know that $s^{\infty}$ is a prefix of $s(C,a)$, but since at every step of the process the length of the sequence strictly increases we must have that $s^{\infty}$ has infinite length, and so $s^{\infty} = s(C,a)$. 
\end{proof}

Note that we must specify that $a \neq 1$, because otherwise $s^{(i)}$ is always $1$, and so the length of the sequence does not grow.

Above we only considered starting sequences which have length one,  but we would like to consider arbitrary starting sequences. Going back to the case of $C = (1 2)$, we can let $s^{(0)}$ be \emph{any} sequence of 1's and 2's that starts with a 2 and the limiting sequence will still be the self-referential sequence.
For example, suppose $s^{(0)} = 2,1,1,1,2$. The rules for generating $s^{(i)}$ give:
\begin{align*}
s^{(0)}&= 2,1,1,1,2\\
s^{(1)} &= 2,2,1,2,1,2,2\\
s^{(2)} &= 2,2,1,1,2,1,1,2,1,1,2,2\\
s^{(3)} &= 2,2,1,1,2,1,2,2,1,2,1,1,2,1,2,2,1,1\\
&\vdots
\end{align*}
We claim that $s^\infty = s(C,a)$, which we state as the following corollary.

\begin{corollary}
Let $C$ be some generating cycle, let $s^{(0)}$ be any sequence of elements of $C$ which begins with some $a \in C$ such that $a \neq 1$. Then $s^{(i)}$ is weakly convergent to $s(C,a)$. 
\end{corollary}
\begin{proof} 
This follows directly from Proposition \ref{iteratetheorem}. That is because $s^{(0)}$ begins with an $a$, at each step of the process $s^{(i)}$ will be a prefix of $s(C,a)$ followed by some terms that disagree with $s(C,a)$. In the limit the prefix of $s(C,a)$ becomes $s(C,a)$ and so $s^{\infty} = s(C,a)$.
\end{proof}
Note that we do not guarantee strong convergence as there are some starting sequences, such as $2222$, which in numerical simulation appears to produce sequences which disagree with the $s( (12), 2)$ on a constant fraction of indices. We would like to be able to produce a general result that distinguishes between strongly and weekly convergent subsequences but it is unclear that there is a nice categorization of them.

@@

The iterative process given in Section~\ref{iterativeprocess} is one way to generate self-referencing sequences. However, for certain generating cycles $C$, there is a simpler way to perform this iteration using a fixed set of substitution rules. Take $C = (1 3)$ as an example, and consider the sequence that starts with a 3:
\begin{align*}
333111333131333111333\dots
\end{align*}
We can use the iterative process from Section~\ref{iterativeprocess}, and suppose that $s^{(0)} = 33$. We get:
\begin{align*}
\{s^{(0)}\} &= 3,3\\
\{s^{(1)}\} &= 3,3,3,1,1,1\\
\{s^{(2)}\} &= 3,3,3,1,1,1,3,3,3,1,3,1\\
\{s^{(3)}\} &= 3,3,3,1,1,1,3,3,3,1,3,1,3,3,3,1,1,1,3,3,3,1,3,3,3,1\\
&\cdots
\end{align*}
It turns out that we can model this evolution with the following substitution rules:
\begin{align*}
\text{Rule }1: 3,3 &\to 3,3,3,1,1,1 \\
\text{Rule }2: 3,1 &\to 3,3,3,1\\
\text{Rule }3: 1,1 &\to 3,1
\end{align*}
This is to be interpreted as follows. Starting from 3,3, the only possible rule to apply is Rule 1, which gives $3,3,3,1,1,1$ at the following iteration. We then apply Rule 1 to $3,3$, Rule 2 to $3,1$ and Rule 3 to $1,1$. Note that at every step, we break up the sequence into chunks of 2 and then simultaneously apply \emph{all} the rules. This will always be possible, since the rules preserve the fact that these sequences have even length. So at the following step, we have 3,3,3,1,1,1,3,3,3,1,3,1, which is $s^{(2)}$. Notice that in general, after the $i$th application of the substitution rules, the sequence is equal to $s^{(i)}$.

We can formulate a set of sufficient conditions for rules that generate the self-referencing sequences.

\textbf{Reader's Note:} The following formalism might be hard to follow, but the general idea of the formalism states that we should do what we did in the $C = (13)$ case. In that example, we wrote down all possible rules that transform strings of a certain length (2), such that each rule was self referential (the input to the rule specifies the block lengths of the output to the rule).

\begin{proposition}
\label{ruleprop}
Suppose $C$ is a generating cycle. Fix a length $l$ that is divisible by $|C|$ as well as some $a \in C$. Then for all possible substrings $x$ (using numbers in $C$) of length $l$, write down the unique rule, if it exists, of the form $x \to y$ that satisfies the following properties:
\begin{enumerate}
\item $x$ is the sequence of the block lengths of $y$.
\item $y$ starts $a$
\item $|y|$ is divisible by $l$.
\end{enumerate}
If any string $x$ does not have a corresponding $y$ that satisfies the above properties, then the rule set does not work. If we can write down a rule for each $x$, then successively applying these rules to a starting sequence length $l$ will give sequences that weakly converge to $s(C,a)$.
\end{proposition}

\begin{proof}
Let $s^{(i)}$ be the sequence generated by the iterative process from Section~\ref{iterativeprocess} given some starting sequence $s^{(0)}$ of length $l$. We show by induction on $i$ that applying the above rules to the same starting sequence simulates the iterative process.

The base case is trivially satisfied, since we start with the same starting sequence $s^{(0)}$ in both the iterative process and this substitution process.

For the inductive step, suppose that applying the rules in $i$ times to $s^{(0)}$ gives $s^{(i)}$, the sequence given at the $i$th step of the iterative process. Note that $|s^{(i)}|$ is divisible by $l$ since each iteration applies the rule to a segment of length $l$ and by condition 3, it writes a sequence of length a multiple of $l$. So we can partition $s^{(i)}$ uniquely into disjoint chunks of size $l$.

In the iterative process, each subsequence of size $l$ ``generates" $l$ blocks at the next iteration, and since $|C|$ divides $l$, the block referenced by each of these subsequences will start with the same number the subsequence started with. Additionally, since all the rules begin with $a$, this number will be $a$.

So applying the rules to each chunk is equivalent to applying the iterative process, where we group elements into subsequences of size $l$. Therefore, $s^{(i+1)}$ is the result of applying the rules to $s^{(i)}$.
\end{proof}

Given Proposition~\ref{ruleprop} we can generate rules a large class of generating cycles $C$. Take the following example, where $C = (2 4)$, and the rules are
\begin{align*}
\text{Rule }1: 2,2 &\to 2, 2, 4, 4 \\
\text{Rule }2: 2,4 &\to 2, 2, 4,4,4,4\\
\text{Rule }3: 4,2 &\to 2,2,2,2,4,4\\
\text{Rule }4: 4,4 &\to 2, 2, 2, 2, 4, 4, 4, 4.
\end{align*}
These rules satisfy the conditions of Proposition~\ref{ruleprop}, and so successive application of these rules will generate a self-referencing sequence which is weakly convergent to $s(C, 2)$.

Unfortunately, this proposition will not help us resolve the density problem for $C = (12)$.

@@

\begin{proof}
Suppose $l$ is a positive integer. Then, by our definition of $l$ in Proposition~\ref{ruleprop}, $l$ must be even. Let $s$ be the string with one 2 and $l-1$ 1's. Suppose $x \rightarrow y$ is the rule for $s$, so $x = s$. Then by condition 1 of Proposition~\ref{ruleprop}, $|y| = l+1$, so $|y|$ is odd, but then it is not divisible by $l$, as condition 3 requires. This is a contradiction.
\end{proof}

@@

We consider the idea of density, which is the proportion of a self-referential sequence that is equal to a certain number. We build off of the substitution rules discovered in the previous section to study the densities in sequences with generating cycle $C = (1 3)$ (note that the sequence that starts with a 1 differs by only one number from the sequence that starts with a 3, so we do not need to consider them separately). While we cannot prove that a limiting density exists, we give a numerical calculation of what the density is \emph{if it exists}. We use a general technique that takes advantage of the fact that substitution rules when $C = (1 3)$, as shown in Section~\ref{substitutionrules}.
